Moisture absorption and drying alter nonadditive litter flammability in a mixed conifer forest

### 2020-06-23 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Edits for R4 and response to R3

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible
that similar nonadditive effects may influence litter moisture content and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with a single study finding increasing
nonadditivity with moisture \citep{Blauw+Wensink+etal-2015}, but this previous
work did not include estimation of mixture effect on the dry-down process and
had an unclear relationship to fire behavior in the field. In mixed conifer
forest in California, where the fire regime is characterized by surface fires
fueled by plant litter, overstory tree composition drives variation in local
fire behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns
\citep{Schwilk+Caprio-2011}, but current estimates of litter flammability in
mixed conifer forests are based largely on burning dry litter.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider ``litter'' the top layer of leaves and small
twigs less than 1\,cm diameter that have fallen in the previous year (mostly
undecomposed or lightly decomposed litter representing fuels classified as
1-hour and 10-hour fuels). Leaf litter was collected in the summer (mid-June to
mid-July) in 2012 from 21 separate sites across the study area. We collected
from a minimum of four populations per species with each population within a
species was 9--32\,km distant from its neighbors. An exception to this was
\emph{A. magnifica} which is restricted to the higher elevations and could only
be collected from one area due to designated wilderness boundaries and
logistical constraints. Within that area, \emph{A. magnifica} litter was
collected from over 10 different individuals. For the remaining species, the
collection involved 2 -- 4 individual trees at each site which were separated
from one another by at least 10\,m. The litter was collected 2\,m away from the
trunk to obtain a more uniform sample of leaf and fine branchlet litter,
because bark and heavier branches tend to fall closer to the tree trunk. The
top 2.5 -- 5\,cm of litter were collected by raking and shoveling into large bags.
All litter was oven dried at $60^\circ$C to $< 5$\% fuel moisture by dry
weight, then weighed and divided into moisture trials. The collected litter was
stored in large paper bags and cardboard boxes in a lab storage space. We
minimized handling and moving the bagged litter to avoid mechanical damage and
changes to the particle size distributions. The moisture manipulations and
subsequent burning trials were conducted in April -- August 2015.

@@

Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that
species within a genus tended to behave similarly and that the two scale-leaved
genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
behaved similarly to one another. Therefore, to preserve power we used
taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus})
as our flammability functional groups and treated these as fixed effects in our
models. The effects of fuel moisture and time since wetting on flammability
parameters were modeled using linear models with fuel moisture or time since
wetting and taxon as fixed effects and average vapor pressure deficit in the
burning room during that trial as a nuisance covariate. We used linear mixed
models to predict spread rate, but fuel consumption tends to exhibit threshold
effects with few intermediate values. Therefore, we used mixed generalized
linear models with a beta regression link with the `glmmADMB' package
\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
of moisture content and of time since wetting.

@@

\begin{figure}
  \centering
  \includegraphics[width=16cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
  \caption{}
\end{figure}
%%\clearpage
\begin{figure}
  \captionsetup{labelformat=adja-page}
  \ContinuedFloat
  \caption{Effects of litter drying on fuel moisture and flammability. A) Dry
    down curves for eight species. Lines show best-fit exponential decay curves
    fit to moisture content on a dry mass basis. B) Flame spread rate by time
    since wetting. Lines indicate best fit linear model by taxonomic group.
    Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in
    SI Table 6. C) Percent fuel consumed by time since wetting. Curves indicate
    best fit beta regression lines by taxonomic group. Model ANOVA table in
    Table \ref{tab:consume_anova} and coefficients in SI Table 7. D) Flame
    spread rate by fuel moisture content. Model ANOVA table in Table
    \ref{tab:spreadrate_anova} and coefficients in SI Table 4. E) Percent fuel
    consumed by moisture content. Lines indicate best fit beta regression
    lines. Model deviance table in Table \ref{tab:consume_anova} and
    coefficients are in SI Table 5. For plots in left column, horizontal jitter
    was added to avoid overplotting. for flammability plots, a lack of data at
    earlier time points for a taxon indicates no ignitions occurred at those
    time points.}
 \label{fig:drydown}
\end{figure}
%% \clearpage

>>>>

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible
that similar nonadditive effects may influence litter moisture content and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with a single study finding increasing
nonadditivity with moisture \citep{Blauw+Wensink+etal-2015}. However, this study
did not include estimation of mixture effect on the dry-down process and
had an unclear relationship to fire behavior in the field. In mixed conifer
forest in California, where the fire regime is characterized by surface fires
fueled by plant litter, overstory tree composition drives variation in local
fire behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns
\citep{Schwilk+Caprio-2011}, but current estimates of litter flammability in
mixed conifer forests are based largely on burning dry litter.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Douglas, \emph{Pinus ponderosa} Douglas ex Lawson \& C. Lawson,
\emph{Abies concolor} (Gordon \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murray, \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz (identification and authorities from the Jepson Flora Project 2020). In this study, we consider ``litter'' the top layer of leaves and small
twigs less than 1\,cm diameter that have fallen in the previous year (mostly
undecomposed or lightly decomposed litter representing fuels classified as
1-hour and 10-hour fuels). Leaf litter was collected in the summer (mid-June to
mid-July) in 2012 from 21 separate sites across the study area. We collected
from a minimum of four populations per species with each population within a
species was 9--32\,km distant from its neighbors. An exception to this was
\emph{A. magnifica} which is restricted to the higher elevations and could only
be collected from one area due to designated wilderness boundaries and
logistical constraints. Within that area, \emph{A. magnifica} litter was
collected from over 10 different individuals. For the remaining species, the
collection involved 2 -- 4 individual trees at each site which were separated
from one another by at least 10\,m. The litter was collected 2\,m away from the
trunk to obtain a more uniform sample of leaf and fine branchlet litter,
because bark and heavier branches tend to fall closer to the tree trunk. The
top 2.5 -- 5\,cm of litter were collected by raking and shoveling into large bags.
All litter was oven dried at $60^\circ$C to $< 5$\% fuel moisture by dry
weight, then weighed and divided into moisture trials. The collected litter was
stored in large paper bags and cardboard boxes in a lab storage space. We
minimized handling and moving the bagged litter to avoid mechanical damage and
changes to the particle size distributions. The moisture manipulations and
subsequent burning trials were conducted in April -- August 2015.

@@

Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that
species within a genus tended to behave similarly and that the two scale-leaved
genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
behaved similarly to one another. Therefore, to preserve power we used
taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus})
as our flammability functional groups and treated these as fixed effects in our
models (Table \ref{tab:species}). The effects of fuel moisture and time since wetting on flammability
parameters were modeled using linear models with fuel moisture or time since
wetting and taxon as fixed effects and average vapor pressure deficit in the
burning room during that trial as a nuisance covariate. We used linear mixed
models to predict spread rate, but fuel consumption tends to exhibit threshold
effects with few intermediate values. Therefore, we used mixed generalized
linear models with a beta regression link with the `glmmADMB' package
\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
of moisture content and of time since wetting.

\begin{table}
  \caption{Species included in this study and the taxonomic groups used as fixed effects in statistical models. Names and authorities according to \citet{Jepson:2020}.
  \label{tab:species}}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{lll}
  \toprule
  Species & Authority & Taxonomic group \\
  \midrule
    \emph{Pinus jeffreyi}    & Grev. \& Balf.  & \emph{Pinus}\\
    \emph{Pinus lambertiana} & Douglas          & \emph{Pinus} \\
    \emph{Pinus ponderosa}   &  Douglas ex Lawson \& C. Lawson & \emph{Pinus} \\
    \emph{Abies concolor}    & (Gordon \& Glend.) Lindl. ex Hildebr. & \emph{Abies} \\
    \emph{Abies magnifica}   & A. Murray   & \emph{Abies} \\
    \emph{Calocedrus decurrens} & (Torr.) Florin & Cuppressaceae \\
    \emph{Sequoiadendron giganteum} & (Lindl.) J. Buchholz & Cuppressaceae \\
    \emph{Quercus kelloggii}    & Newb. & \emph{Quercus}\\ 
  \bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
  \centering
  \includegraphics[width=16cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
  \caption{}
\end{figure}
%%\clearpage
\begin{figure}
  \captionsetup{labelformat=adja-page}
  \ContinuedFloat
  \caption{Effects of litter drying on fuel moisture and flammability. A) Dry
    down curves for eight species. Lines show best-fit exponential decay curves
    fit to moisture content on a dry mass basis. The lines for \emph{P.
      jeffreyi} and for \emph{P. ponderosa} overlap and appear to be a single
    line. B) Flame spread rate by time since wetting. Lines indicate best fit
    linear model by taxonomic group (Table \ref{tab:species}). Model ANOVA
    table is in Table \ref{tab:spreadrate_anova} and coefficients are in SI Table 6.
    C) Percent fuel consumed by time since wetting. Curves indicate best fit
    beta regression by taxonomic group, except that \emph{P. lambertiana}
    is fit separately from \emph{P. jeffrey} and \emph{P. ponderosa} as
    explained in the text (Table \ref{tab:species}). Model ANOVA table is in Table
    \ref{tab:consume_anova} and coefficients are in SI Table 7. D) Flame spread
    rate by fuel moisture content. Lines indicate best fit linear model by
    taxonomic group (Table \ref{tab:species}). Model ANOVA table is in Table
    \ref{tab:spreadrate_anova} and coefficients are in SI Table 4. E) Percent fuel
    consumed by moisture content. Lines indicate best fit beta regression
    lines. Curves indicate best fit linear model by taxonomic group (Table
    \ref{tab:species}). Model deviance is in Table \ref{tab:consume_anova}
    and coefficients are in SI Table 5. For plots in left column, horizontal
    jitter was added to avoid overplotting. For flammability plots, a lack of
    data at earlier time points for a taxon indicates no ignitions occurred at
    those time points.}
 \label{fig:drydown}
\end{figure}
%% \clearpage
